\chapter*{Abstract}

%\begin{Notes}
%- Make corrections.\\
%- Check portuguese version.\\
%\end{Notes}

The advent of next generation sequencing methods has revolutionized the field of
molecular biology in the past few years. Nowadays, we are able to produce
enormous amounts of biological information, both quickly and at low cost. As
such, tools have to evolve accordingly, in order to cope with such large volumes
of information. In this report we discuss the usage of computer tools capable of
conducting gene expression profiling based on information obtained through RNA
Sequencing techniques, applied to a specific set of biological problems. In
particular, we present the idealization process and implementation details of a
web platform capable of addressing these problems, as well as the actual
platform prototype. The prototype's functionality is showcased with a real case
study, produced in collaboration with biology researchers. This report also
includes a literature review, covering both the biological and technical aspects
of the work, with special emphasis in machine learning techniques applied to
data mining tasks. Lastly, we review the work done and results obtained so far
and outline the possible future of the web platform.

\chapter*{Resumo}

\begin{otherlanguage}{portuguese}
O advento das técnicas de sequenciação de nova geração revolucionou o campo da
biologia molecular nos últimos anos. Hoje em dia somos capazes de produzir
enormes quantidades de informação biológica rapidamente e a baixo custo. Assim
sendo, as ferramentas devem também evoluir, a fim de lidarem com estas extensas
quantidades de informação. Neste relatório discutimos o uso de ferramentas
informáticas capazes de analisar perfis de expressão génica com base em
informação obtida através de técnicas de \textit{\rnaseq{}}, aplicadas a um
conjunto específico de problemas biológicos. Em particular, apresentamos o
processo de idealização e os detalhes de implementação de uma plataforma
\textit{web} capaz de resolver estes problemas, assim como o protótipo funcional
dessa plataforma. As funcionalidades deste protótipo são demonstradas através de
um caso de estudo real, produzido em colaboração com investigadores da área da
biologia. Este relatório inclui também uma revisão da literatura, cobrindo os
aspetos biológicos e técnicos deste trabalho, com um ênfase especial em técnicas
de aprendizagem máquina aplicadas a tarefas de \textit{data mining}. Por fim,
revemos todo o trabalho efetuado e os resultados obtidos até ao momento e
delineamos as possibilidades futuras para a plataforma \textit{web}.
\end{otherlanguage}
