\chapter{Conclusions} \label{chap:conclusions}

\section*{}

%\begin{Notes}
%- Check PDIS conclusion.\\
%- Talk about finishing iRAP's web integration and further exploration of its
%results.\\
%- Talk about full automated integration between both tools.\\
%\end{Notes}

In this thesis we developed a prototype of an integrated platform for RNA-Seq
data analysis. This platform is able to perform the entire analysis process,
from the sequencing reads, to grouping genes by their RBPs. This means that is
able to perform read alignment, quantification and differential expression
analysis tasks, as well as data set enrichment, RBP identification and
clustering analysis of genes and proteins. Through two case studies we showed
the developed prototype in use, and assessed its correctness and efficiency.
Below we will share our thoughts on the fulfilment of the objectives of this
thesis, and present some possibilities for any future development in our
solution.

During this thesis we were advised by molecular biology experts from IBMC. The
data set used in the second case study was provided by IBMC experts and the
results obtained are currently being experimentally verified in their
laboratory.

\section{Objective Fulfilment}

Our objectives, in terms of studying the problem at hand and developing a
solution to it, were completely fulfilled. The proposed solution corresponds to
all of our expectations. Both analysis pipelines were completely prototyped and
tested, being that PBS Finder was tested in both development and production
environments. However, it was not possible to integrate the RNA-seq data
analysis pipeline with PBS Finder's analysis pipeline, due to time constraints.
Similarly, it was also not possible to provide the RNA-Seq analysis pipeline's
functionality through the web interface. As such, our objective of prototyping
and testing the complete integrated system could not be completely achieved.

\section{Future Work}

A natural continuation of the proposed work would be to finish the
implementation and integration of the RNA-Seq data analysis pipeline. This would
allow our solution to work as designed, integrating the complete analysis
pipeline, from sequencing data to gene clustering and result visualization.
Furthermore, it would be useful to study the developed tools in terms of
performance, under large volumes of information and requests. Whilst the tools
were developed taking in consideration their performance, making them available
in a large scale would require another kind of infrastructure.
